---
title: "Soil Organic Carbon Prediction from VisNIR Spectroscopy"
subtitle: "Random Forest Prototype · Report 01 · BMP Project"
date: "March 2026"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: true
    number_sections: true
    df_print: paged
    fig_width: 9.5
    fig_height: 5.2
    self_contained: true
---

```{css echo=FALSE}
/* ── Typography ───────────────────────────────────────────────────────── */
body { font-size: 15px; color: #2c3e50; line-height: 1.75; }
h1, h2, h3 { color: #2c3e50; font-weight: 600; }
h1.title  { font-size: 2.1em; color: #1a252f; }
h3.subtitle { font-size: 1.15em; color: #576574; font-weight: 400; margin-top: 0; }

/* ── Callout boxes ────────────────────────────────────────────────────── */
.callout-blue {
  background: #ebf5fb; border-left: 5px solid #2980b9;
  padding: 14px 18px; border-radius: 4px; margin: 18px 0;
}
.callout-green {
  background: #eafaf1; border-left: 5px solid #27ae60;
  padding: 14px 18px; border-radius: 4px; margin: 18px 0;
}
.callout-orange {
  background: #fef9e7; border-left: 5px solid #e67e22;
  padding: 14px 18px; border-radius: 4px; margin: 18px 0;
}
.callout-red {
  background: #fdedec; border-left: 5px solid #e74c3c;
  padding: 14px 18px; border-radius: 4px; margin: 18px 0;
}

/* ── Author block ─────────────────────────────────────────────────────── */
.author-block {
  background: #f8f9fa; border: 1px solid #dee2e6;
  border-radius: 6px; padding: 16px 22px; margin: 16px 0 26px 0;
  display: flex; align-items: flex-start; gap: 20px;
}
.author-info h4 { margin: 0 0 4px 0; font-size: 1.1em; }
.author-info p  { margin: 0; color: #576574; font-size: 0.92em; }
.author-links a {
  display: inline-block; margin: 3px 4px 0 0;
  padding: 3px 10px; border-radius: 12px; font-size: 0.82em;
  text-decoration: none; color: white; font-weight: 500;
}
.orcid-btn   { background: #A6CE39; }
.github-btn  { background: #181717; }
.rg-btn      { background: #00CCBB; }
.li-btn      { background: #0A66C2; }
.x-btn       { background: #000000; }

/* ── Tables ───────────────────────────────────────────────────────────── */
table { width: 100%; border-collapse: collapse; margin: 12px 0 20px 0; font-size: 0.93em; }
thead tr { background: #2c3e50; color: white; }
th { padding: 9px 13px; text-align: left; }
td { padding: 8px 13px; border-bottom: 1px solid #ecf0f1; }
tr:nth-child(even) { background: #f8f9fa; }
tr.best-row { background: #eafaf1 !important; font-weight: 600; }

/* ── Metric highlight boxes ───────────────────────────────────────────── */
.metric-grid {
  display: grid; grid-template-columns: repeat(auto-fit, minmax(130px, 1fr));
  gap: 12px; margin: 18px 0;
}
.metric-box {
  background: #f8f9fa; border-radius: 6px; padding: 14px 10px;
  text-align: center; border-top: 4px solid #2980b9;
}
.metric-box.best { border-top-color: #27ae60; background: #eafaf1; }
.metric-val  { font-size: 1.7em; font-weight: 700; color: #2c3e50; line-height: 1; }
.metric-name { font-size: 0.78em; color: #7f8c8d; margin-top: 4px; text-transform: uppercase; letter-spacing: 0.04em; }
.metric-sd   { font-size: 0.75em; color: #95a5a6; }

/* ── Figure captions ──────────────────────────────────────────────────── */
p.caption { color: #7f8c8d; font-size: 0.88em; text-align: center; margin-top: -8px; }

/* ── Code ─────────────────────────────────────────────────────────────── */
pre { background: #f4f6f7; border-radius: 4px; font-size: 0.85em; }
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  dpi     = 150,
  cache   = FALSE
)

# Set working directory to project root for all chunks
knitr::opts_knit$set(root.dir = normalizePath("../../"))
```

```{r libs, include=FALSE}
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(readr)
  library(ggplot2)
  library(ggpubr)
  library(viridis)
  library(scales)
})

# ── Source metrics utilities (for fmt_ms helper) ────────────────────────
source("R/utils_metrics.R")

# ── Helper: formatted mean ± SD string ─────────────────────────────────
fmt_ms <- function(m, s, digits = 2)
  sprintf("%s ± %s", round(m, digits), round(s, digits))

# ── Colour palette (three treatments) ──────────────────────────────────
prep_cols <- c(
  sg_smooth = "#2980b9",
  sg_deriv1 = "#27ae60",
  snv       = "#e67e22"
)
prep_labels <- c(
  sg_smooth = "SG Smooth",
  sg_deriv1 = "SG 1st Derivative",
  snv       = "SNV"
)
```

```{r load-report-data, include=FALSE}
# ── Load compact pre-processed report dataset ───────────────────────────
# Generated by prepare_report_data.R — run that script first if missing.
rd <- readRDS("reports/report_01_soc_ossl_prototype/report_data.rds")

X_raw         <- rd$X_raw
spec_list     <- rd$spec_list
y_orig        <- rd$y_orig
uuid_id       <- rd$uuid_id
wave_nm       <- rd$wave_nm
n_wave        <- rd$n_wave
n_rows_total  <- rd$n_rows_total
n_uuid_total  <- rd$n_uuid_total
avg_scans     <- rd$avg_scans
n_joined      <- rd$n_joined
n_uuid_joined <- rd$n_uuid_joined
n_final       <- rd$n_final
n_uuid_final  <- rd$n_uuid_final
WAVE_MIN      <- rd$WAVE_MIN
WAVE_MAX      <- rd$WAVE_MAX
OSSL_FRACTION <- rd$OSSL_FRACTION
N_MAX_SAMPLES <- rd$N_MAX_SAMPLES
SG_WINDOW     <- rd$SG_WINDOW
SG_POLY       <- rd$SG_POLY
SEED          <- rd$SEED
rm(rd)
```

```{r load-metrics, include=FALSE}
# ── Load CV metrics (from completed pipeline run) ───────────────────────
m_log <- readr::read_csv(
  "output/metrics/cv_metrics_oc_usda.c729_w.pct_log.csv",
  show_col_types = FALSE
)
m_raw <- readr::read_csv(
  "output/metrics/cv_metrics_oc_usda.c729_w.pct_raw.csv",
  show_col_types = FALSE
)
m_comp <- readr::read_csv(
  "output/metrics/log_comparison_oc_usda.c729_w.pct.csv",
  show_col_types = FALSE
)

# Best model overall (RAW | sg_deriv1)
best_row <- m_raw %>% dplyr::filter(preprocessing == "sg_deriv1")
```

---

<div class="author-block">
<div class="author-info">
<h4>Hugo Machado Rodrigues</h4>
<p>University of Florida &mdash; BMP Project &nbsp;|&nbsp; Soil Spectroscopy &amp; Machine Learning</p>
<div class="author-links">
<a class="orcid-btn"  href="https://orcid.org/0000-0002-8070-8126"                        target="_blank">ORCID</a>
<a class="github-btn" href="https://github.com/HugoMachadoRodrigues"                       target="_blank">GitHub</a>
<a class="rg-btn"     href="https://www.researchgate.net/profile/Hugo-Rodrigues-12"        target="_blank">ResearchGate</a>
<a class="li-btn"     href="https://www.linkedin.com/in/hugo-rodrigues-52b535119/"         target="_blank">LinkedIn</a>
<a class="x-btn"      href="https://x.com/Hugo_MRodrigues"                                 target="_blank">X / Twitter</a>
</div>
</div>
</div>

---

<div class="callout-blue">
**Executive Summary.** A Random Forest (RF) prototype was developed to predict soil organic carbon (SOC, %) from VisNIR reflectance spectra (400–2450 nm, ASD FieldSpec 4) using the Open Soil Spectral Library (OSSL v1.2). A sub-sample of **`r n_final` observations** from `r n_uuid_final` unique soil layers was used. Three spectral preprocessing treatments (SG smooth, SG 1st derivative, SNV) were evaluated under two response-scale strategies (log₁p-transformed and original scale) via a full 3D hyperparameter grid search (192 combinations) and **10-fold × 3-repeat grouped cross-validation** (30 resamples). The best model — **SG 1st derivative + original scale** — achieved R² = **`r round(best_row$r2, 2)`**, RMSE = **`r round(best_row$rmse, 2)` %**, RPD = **`r round(best_row$rpd, 2)`**, and CCC = **`r round(best_row$ccc, 2)`**, indicating good-to-excellent quantitative predictive performance.
</div>

---

# Study Overview

This report documents the first prototype iteration of the **BMP Soil Spectral Modeling** pipeline — a reproducible, parallelised Random Forest workflow linking VisNIR reflectance spectra to soil laboratory measurements. The prototype uses OSSL data to validate the pipeline before the Years 1–5 BMP project dataset becomes available.

The pipeline is inspired by and extended from **Clingensmith & Grunwald (2022)**, who demonstrated that RF models can effectively predict a wide range of soil properties across the continental United States using VisNIR spectroscopy. Key extensions relative to that work include:

- Full 3D hyperparameter grid search (versus default `ranger` settings)
- Grouped cross-validation folds by layer UUID (leakage prevention)
- Dual log-mode evaluation with side-by-side metric comparison
- MSD decomposition for bias diagnostics (Gauch et al. 2003)
- Lin's CCC and RPIQ in addition to standard metrics

---

# Dataset Description

## Data source: OSSL v1.2

All spectral and laboratory data were obtained from the **Open Soil Spectral Library (OSSL), version 1.2** (Safanelli et al. 2023), a globally distributed open repository of soil spectra paired with laboratory analyses. OSSL data are licensed under [CC BY 4.0](https://creativecommons.org/licenses/by/4.0/).

| Field | Value |
|---|---|
| Library | Open Soil Spectral Library (OSSL) |
| Version | 1.2 |
| Spectral file | `ossl_visnir_L0_v1.2.csv.gz` |
| Laboratory file | `ossl_soillab_L1_v1.2.csv.gz` |
| Sensor | ASD FieldSpec 4 (VisNIR) |
| URL | [soilspectroscopy.org](https://soilspectroscopy.org) |

## Wavelength filtering and spectral completeness

The VisNIR spectral data span 350–2500 nm at 1-nm resolution. The following preprocessing decisions were applied before any modeling:

1. **Edge trimming:** Wavelengths below 400 nm and above 2450 nm were excluded to remove edge-of-range noise common to FieldSpec instruments (Clingensmith & Grunwald 2022). The retained range comprises **`r n_wave` wavelengths** (400–2450 nm).

2. **Completeness filter:** Only observations with non-missing reflectance across all `r n_wave` wavelengths were retained. This effectively restricts the dataset to ASD FieldSpec records, which provide continuous coverage across the full range.

3. **Lab data join:** The spectral dataset was inner-joined with laboratory SOC measurements, retaining only observations with a valid measured SOC value.

## Sample structure and spectral replicates

```{r sample-structure-table}
data.frame(
  Stage = c(
    "Total complete spectral observations (400–2450 nm)",
    "  of which: unique soil layers (layer UUID)",
    "  Average scans per layer",
    "After inner join with SOC lab data",
    paste0("After random ", round(OSSL_FRACTION * 100), "% sampling + N ≤ ", N_MAX_SAMPLES, " cap"),
    "  Unique soil layers in modelling dataset"
  ),
  N = c(
    format(n_rows_total, big.mark = ","),
    format(n_uuid_total, big.mark = ","),
    sprintf("%.2f", avg_scans),
    format(n_joined, big.mark = ","),
    format(n_final, big.mark = ","),
    format(n_uuid_final, big.mark = ",")
  )
) |>
  knitr::kable(col.names = c("Data stage", "Count"), align = c("l","r"))
```

<div class="callout-orange">
**Spectral replicates and leakage prevention.** Multiple spectral scans (`r sprintf("%.2f", avg_scans)` on average) may correspond to the same soil layer and therefore share a single laboratory measurement. These replicates were **retained without averaging** to preserve all available spectral information. However, to prevent data leakage — i.e., the same soil layer appearing in both the training and validation partition of the same fold — cross-validation fold assignment was performed at the **layer UUID level** (Kaufman & Rosset 2012). All scans belonging to the same layer are always assigned to the same fold.
</div>

---

# Target Variable: Soil Organic Carbon

## Summary statistics

```{r soc-stats}
q <- quantile(y_orig, c(0.25, 0.5, 0.75), na.rm = TRUE)
n_zeros <- sum(y_orig == 0, na.rm = TRUE)

data.frame(
  Statistic = c("N", "Minimum", "Q1 (25th pctile)", "Median", "Q3 (75th pctile)",
                "Maximum", "Mean", "SD", "IQR", "Skewness", "Zeros / near-zero"),
  Value = c(
    length(y_orig),
    round(min(y_orig, na.rm = TRUE), 2),
    round(q[1], 2),
    round(q[2], 2),
    round(q[3], 2),
    round(max(y_orig, na.rm = TRUE), 2),
    round(mean(y_orig, na.rm = TRUE), 2),
    round(sd(y_orig, na.rm = TRUE), 2),
    round(q[3] - q[1], 2),
    round(moments::skewness(y_orig, na.rm = TRUE), 2),
    n_zeros
  )
) |>
  knitr::kable(col.names = c("Statistic", "SOC (%)"), align = c("l","r"))
```

## Distribution: raw and log-transformed

SOC distributions are typically right-skewed — a small number of samples with very high organic matter content pull the tail rightward. The log₁p transformation (`log(SOC + 1)`) compresses the upper tail, producing a more symmetric distribution and reducing the influence of extreme values on model training.

```{r soc-histogram, fig.height=4.5, fig.cap="**Figure 1.** Distribution of soil organic carbon (SOC, %) before (left) and after log₁p transformation (right). The dashed red line marks the mean. The log-transformed distribution is substantially more symmetric, reducing skewness and potential leverage of high-SOC samples."}
skew_raw <- round(moments::skewness(y_orig, na.rm = TRUE), 2)
skew_log <- round(moments::skewness(log1p(y_orig), na.rm = TRUE), 2)

df_hist <- data.frame(raw = y_orig, log = log1p(y_orig))

p1 <- ggplot(df_hist, aes(x = raw)) +
  geom_histogram(bins = 40, fill = "#2980b9", colour = "white", linewidth = 0.3) +
  geom_vline(xintercept = mean(y_orig, na.rm = TRUE),
             linetype = "dashed", colour = "#e74c3c", linewidth = 0.8) +
  annotate("text", x = mean(y_orig, na.rm = TRUE) * 1.08,
           y = Inf, vjust = 1.5, hjust = 0,
           label = sprintf("mean = %.2f%%", mean(y_orig, na.rm = TRUE)),
           colour = "#e74c3c", size = 3.4) +
  labs(
    title    = "Raw SOC",
    subtitle = sprintf("Skewness = %s  |  n = %d", skew_raw, length(y_orig)),
    x        = "SOC (%)",
    y        = "Count"
  ) +
  theme_bw(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))

p2 <- ggplot(df_hist, aes(x = log)) +
  geom_histogram(bins = 40, fill = "#27ae60", colour = "white", linewidth = 0.3) +
  geom_vline(xintercept = mean(log1p(y_orig), na.rm = TRUE),
             linetype = "dashed", colour = "#e74c3c", linewidth = 0.8) +
  annotate("text", x = mean(log1p(y_orig), na.rm = TRUE) * 1.04,
           y = Inf, vjust = 1.5, hjust = 0,
           label = sprintf("mean = %.2f", mean(log1p(y_orig), na.rm = TRUE)),
           colour = "#e74c3c", size = 3.4) +
  labs(
    title    = "log₁p(SOC)",
    subtitle = sprintf("Skewness = %s  |  n = %d", skew_log, length(y_orig)),
    x        = "log₁p(SOC)",
    y        = "Count"
  ) +
  theme_bw(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))

ggpubr::ggarrange(p1, p2, ncol = 2)
```

<div class="callout-blue">
**Interpretation.** The raw SOC distribution shows skewness of **`r skew_raw`**, characteristic of right-skewed soil property data. After log₁p transformation, skewness drops to **`r skew_log`**, substantially closer to symmetric. Note, however, that all model performance metrics are computed on the **back-transformed (original %) scale** to ensure interpretability regardless of the transformation used during training.
</div>

---

# Spectral Data

## Sample spectra: raw reflectance

```{r spectra-raw, fig.height=5, fig.cap="**Figure 2.** Random sample of 80 raw reflectance spectra (400–2450 nm, ASD FieldSpec 4), coloured by SOC content (%). Low-SOC soils (yellow) tend to have higher reflectance than high-SOC soils (dark purple), consistent with the well-established negative relationship between organic matter and soil reflectance across the NIR region."}
set.seed(42L)
idx_s <- sample(nrow(X_raw), min(80L, nrow(X_raw)))

long_raw <- data.frame(
  id       = rep(seq_along(idx_s), each = n_wave),
  wl       = rep(wave_nm, times = length(idx_s)),
  ref      = as.vector(t(X_raw[idx_s, ])),
  soc      = rep(y_orig[idx_s], each = n_wave)
)

ggplot(long_raw, aes(x = wl, y = ref, group = id, colour = soc)) +
  geom_line(alpha = 0.55, linewidth = 0.35) +
  scale_colour_viridis_c(
    name   = "SOC (%)",
    option = "plasma",
    trans  = "log1p",
    breaks = c(0.5, 2, 5, 15, 40),
    labels = c("0.5", "2", "5", "15", "40")
  ) +
  scale_x_continuous(
    breaks = seq(500, 2400, by = 250),
    expand = expansion(mult = 0.01)
  ) +
  scale_y_continuous(labels = scales::label_number(accuracy = 0.01)) +
  labs(
    title    = "Raw VisNIR Reflectance Spectra (sample of 80)",
    subtitle = sprintf("ASD FieldSpec 4 · %d–%d nm · %d wavelengths", WAVE_MIN, WAVE_MAX, n_wave),
    x        = "Wavelength (nm)",
    y        = "Reflectance"
  ) +
  theme_bw(base_size = 12) +
  theme(
    legend.position   = "right",
    plot.title        = element_text(face = "bold"),
    panel.grid.minor  = element_blank()
  )
```

## Preprocessing treatments

All three treatments share a **common SG-smoothed base** (Rinnan et al. 2009): the SG smooth (window = `r SG_WINDOW`, poly = `r SG_POLY`) is applied once to raw reflectance, and the 1st-derivative and SNV treatments are derived from that smoothed output. This ensures performance differences are attributable to the transformation step, not to differences in noise reduction.

```{r spectra-treatments, fig.height=7, fig.cap="**Figure 3.** The same random sample of spectra shown under the three preprocessing treatments. **SG Smooth** (top): smoothed reflectance retaining the raw spectral shape. **SG 1st Derivative** (middle): highlights spectral slope changes; additive baseline offsets are removed. **SNV** (bottom): each spectrum is mean-centred and scaled to unit variance; multiplicative scatter is corrected. All treatments computed on the same 80-spectrum sample."}
plots_prep <- lapply(names(prep_labels), function(nm) {
  mat  <- spec_list[[nm]][idx_s, ]
  long <- data.frame(
    id  = rep(seq_along(idx_s), each = ncol(mat)),
    wl  = rep(wave_nm[seq_len(ncol(mat))], times = length(idx_s)),
    ref = as.vector(t(mat)),
    soc = rep(y_orig[idx_s], each = ncol(mat))
  )
  ggplot(long, aes(x = wl, y = ref, group = id, colour = soc)) +
    geom_line(alpha = 0.5, linewidth = 0.3) +
    scale_colour_viridis_c(
      name   = "SOC (%)",
      option = "plasma",
      trans  = "log1p",
      breaks = c(0.5, 2, 5, 15, 40),
      labels = c("0.5", "2", "5", "15", "40"),
      guide  = guide_colourbar(barheight = 4)
    ) +
    scale_x_continuous(
      breaks = seq(500, 2400, by = 250),
      expand = expansion(mult = 0.01)
    ) +
    labs(
      title = prep_labels[[nm]],
      x     = "Wavelength (nm)",
      y     = ""
    ) +
    theme_bw(base_size = 11) +
    theme(
      plot.title       = element_text(face = "bold", colour = prep_cols[[nm]]),
      panel.grid.minor = element_blank(),
      legend.position  = "right"
    )
})

ggpubr::ggarrange(
  plotlist = plots_prep,
  ncol = 1, nrow = 3,
  common.legend = TRUE, legend = "right",
  labels = c("A", "B", "C"), font.label = list(size = 12, face = "bold")
)
```

---

# Modeling Pipeline

## Design overview

```
OSSL VisNIR (350–2500 nm, ASD FieldSpec 4)
         │
         ▼
  Edge trimming (400–2450 nm) + completeness filter
         │
         ▼
  Join with SOC lab data → stratified random sample (n = 1,000)
         │
         ▼
  Raw reflectance
         │
         ▼
  SG Smooth [shared base] ──────────────────► Treatment 1: sg_smooth
         │
         ├──► SG 1st Derivative ────────────► Treatment 2: sg_deriv1
         │
         └──► SNV ──────────────────────────► Treatment 3: snv
                                                          │
                                  ┌───────────────────────┤
                                  │  For each log mode:   │
                                  │  log1p(y) / y_orig    │
                                  └───────────────────────┘
                                                          │
                                  Full 3D grid search (OOB RMSE, parallel)
                                  ntree [500–2000, Δ100] × mtry [p/9, p/6, p/3, p]
                                  × min.node.size [3, 5, 10] = 192 combinations
                                                          │
                                  10-fold × 3-repeat stratified grouped CV
                                  Folds assigned by layer UUID (leakage prevention)
                                                          │
                                  Final model on full dataset
                                                          │
                            Metrics · Figures · Saved models (.rds)
```

## Hyperparameter search space

| Parameter | Candidates | N values |
|---|---|:---:|
| `num.trees` | 500, 600, …, 2000 (step 100) | 16 |
| `mtry` | ⌊p/9⌋, ⌊p/6⌋, ⌊p/3⌋, p | 4 |
| `min.node.size` | 3, 5, 10 | 3 |
| **Total combinations** | **16 × 4 × 3** | **192** |

All 192 models were trained using OOB predictions only (`importance = "none"`) and run simultaneously across all available CPU cores via `foreach %dopar%` with a `doSNOW` progress-bar backend. The combination with the lowest OOB RMSE was selected as the optimal hyperparameter set for each preprocessing × log-mode combination (Breiman 2001; Liaw & Wiener 2002; Probst et al. 2019).

## Cross-validation design

| Setting | Value |
|---|---|
| Strategy | k-fold repeated cross-validation |
| Folds (k) | 10 |
| Repeats | 3 |
| Total resamples | 30 |
| Fold assignment | Grouped by layer UUID (Kaufman & Rosset 2012) |
| Stratification | Quantile bins of group-mean SOC |
| Response metrics | Computed on original % scale after back-transformation |
| Summary | Mean ± SD across 30 resamples |

---

# Cross-Validation Results

## Log₁p-transformed scale (log mode)

In log mode, the RF was trained on log₁p(SOC). Predictions were back-transformed with `expm1()` before computing all metrics.

```{r table-log}
m_log_disp <- m_log %>%
  dplyr::mutate(
    Preprocessing = prep_labels[preprocessing],
    `R²`          = fmt_ms(r2,   r2_sd),
    `RMSE (%)`    = fmt_ms(rmse, rmse_sd),
    `MAE (%)`     = fmt_ms(mae,  mae_sd),
    `ME (%)`      = fmt_ms(me,   me_sd),
    `RPD`         = fmt_ms(rpd,  rpd_sd),
    `RPIQ`        = fmt_ms(rpiq, rpiq_sd),
    `CCC`         = fmt_ms(ccc,  ccc_sd)
  ) %>%
  dplyr::select(Preprocessing, `R²`, `RMSE (%)`, `MAE (%)`, `ME (%)`, RPD, RPIQ, CCC)

knitr::kable(
  m_log_disp,
  caption = "**Table 1.** CV performance metrics under log₁p transformation (mean ± SD across 30 resamples). All metrics computed on the back-transformed SOC (%) scale.",
  align   = c("l", rep("r", 7))
)
```

## Original scale (no-log mode)

In no-log mode, the RF was trained and evaluated directly on SOC (%) without transformation.

```{r table-raw}
m_raw_disp <- m_raw %>%
  dplyr::mutate(
    Preprocessing = prep_labels[preprocessing],
    `R²`          = fmt_ms(r2,   r2_sd),
    `RMSE (%)`    = fmt_ms(rmse, rmse_sd),
    `MAE (%)`     = fmt_ms(mae,  mae_sd),
    `ME (%)`      = fmt_ms(me,   me_sd),
    `RPD`         = fmt_ms(rpd,  rpd_sd),
    `RPIQ`        = fmt_ms(rpiq, rpiq_sd),
    `CCC`         = fmt_ms(ccc,  ccc_sd)
  ) %>%
  dplyr::select(Preprocessing, `R²`, `RMSE (%)`, `MAE (%)`, `ME (%)`, RPD, RPIQ, CCC)

knitr::kable(
  m_raw_disp,
  caption = "**Table 2.** CV performance metrics on the original SOC (%) scale (mean ± SD across 30 resamples).",
  align   = c("l", rep("r", 7))
)
```

## Best model per log mode — key metrics

```{r best-metrics-boxes, fig.height=1.8, out.width="100%", results='asis'}
# Best per mode
best_log <- m_log %>% dplyr::filter(preprocessing == "sg_deriv1")
best_raw <- m_raw %>% dplyr::filter(preprocessing == "sg_deriv1")

cat('<div style="display:grid; grid-template-columns:1fr 1fr; gap:16px; margin:12px 0 24px 0;">')

# LOG box
cat('<div style="border:1px solid #dee2e6; border-radius:6px; padding:16px; background:#f8f9fa;">')
cat('<p style="font-weight:700; font-size:1em; color:#2980b9; margin:0 0 10px 0;">&#x1F4CA; Best model &mdash; Log mode (SG 1st Derivative)</p>')
cat('<div class="metric-grid">')
for(nm in c("r2","rmse","mae","rpd","ccc")) {
  vals <- list(r2=best_log$r2, rmse=best_log$rmse, mae=best_log$mae, rpd=best_log$rpd, ccc=best_log$ccc)
  sds  <- list(r2=best_log$r2_sd, rmse=best_log$rmse_sd, mae=best_log$mae_sd, rpd=best_log$rpd_sd, ccc=best_log$ccc_sd)
  labels_nm <- c(r2="R²", rmse="RMSE (%)", mae="MAE (%)", rpd="RPD", ccc="CCC")
  cat(sprintf('<div class="metric-box"><div class="metric-val">%s</div><div class="metric-name">%s</div><div class="metric-sd">± %s</div></div>',
              round(vals[[nm]], 2), labels_nm[[nm]], round(sds[[nm]], 2)))
}
cat('</div></div>')

# RAW box
cat('<div style="border:1px solid #dee2e6; border-radius:6px; padding:16px; background:#eafaf1;">')
cat('<p style="font-weight:700; font-size:1em; color:#27ae60; margin:0 0 10px 0;">&#x1F3C6; Best model &mdash; No-log mode (SG 1st Derivative)</p>')
cat('<div class="metric-grid">')
for(nm in c("r2","rmse","mae","rpd","ccc")) {
  vals <- list(r2=best_raw$r2, rmse=best_raw$rmse, mae=best_raw$mae, rpd=best_raw$rpd, ccc=best_raw$ccc)
  sds  <- list(r2=best_raw$r2_sd, rmse=best_raw$rmse_sd, mae=best_raw$mae_sd, rpd=best_raw$rpd_sd, ccc=best_raw$ccc_sd)
  labels_nm <- c(r2="R²", rmse="RMSE (%)", mae="MAE (%)", rpd="RPD", ccc="CCC")
  cat(sprintf('<div class="metric-box best"><div class="metric-val">%s</div><div class="metric-name">%s</div><div class="metric-sd">± %s</div></div>',
              round(vals[[nm]], 2), labels_nm[[nm]], round(sds[[nm]], 2)))
}
cat('</div></div>')
cat('</div>')
```

## Observed vs. predicted — SG 1st Derivative

```{r obs-pred-deriv1, fig.height=4.5, fig.cap="**Figure 4.** Observed vs. predicted SOC (%) for the SG 1st Derivative preprocessing treatment. Left: log mode (back-transformed); right: no-log mode. Each point represents one held-out fold observation across the 30 resamples. The dashed line is the 1:1 reference. The solid blue line is the OLS regression of predicted on observed. Tight agreement with the 1:1 line indicates low bias; the regression slope indicates proportional bias."}
knitr::include_graphics(c(
  "output/figures/obs_pred_oc_usda.c729_w.pct_log_sg_deriv1.png",
  "output/figures/obs_pred_oc_usda.c729_w.pct_raw_sg_deriv1.png"
))
```

## Observed vs. predicted — all treatments (no-log)

```{r obs-pred-all-raw, fig.height=4.5, fig.cap="**Figure 5.** Observed vs. predicted SOC (%) for all three preprocessing treatments under the no-log (original scale) strategy. The SG 1st derivative (centre) shows the tightest agreement with the 1:1 line and the highest R²."}
knitr::include_graphics(c(
  "output/figures/obs_pred_oc_usda.c729_w.pct_raw_sg_smooth.png",
  "output/figures/obs_pred_oc_usda.c729_w.pct_raw_sg_deriv1.png",
  "output/figures/obs_pred_oc_usda.c729_w.pct_raw_snv.png"
))
```

## Preprocessing comparison — RPD and R²

```{r comparison-plots, fig.height=4.0, fig.cap="**Figure 6.** Cross-preprocessing CV comparison for the no-log mode. Left: RPD (Residual Prediction Deviation; higher is better, RPD > 2 = good quantitative). Right: R² (coefficient of determination; mean ± SD across 30 resamples). SG 1st derivative consistently outperforms SG smooth and SNV."}
knitr::include_graphics(c(
  "output/figures/comparison_rpd_oc_usda.c729_w.pct_raw.png",
  "output/figures/comparison_r2_oc_usda.c729_w.pct_raw.png"
))
```

---

# Log Transformation: Does It Help?

```{r log-comparison-table}
comp_disp <- data.frame(
  `Log mode`     = c("log₁p(SOC)", "Original SOC (%)"),
  Preprocessing  = prep_labels[m_comp$preprocessing],
  `R²`           = round(m_comp$r2, 2),
  `RMSE (%)`     = round(m_comp$rmse, 2),
  `MAE (%)`      = round(m_comp$mae, 2),
  `RPD`          = round(m_comp$rpd, 2),
  `RPIQ`         = round(m_comp$rpiq, 2),
  `CCC`          = round(m_comp$ccc, 2),
  check.names    = FALSE
)
knitr::kable(
  comp_disp,
  caption = "**Table 3.** Best model per log mode (both selected sg_deriv1). Metrics computed on original SOC (%) scale after back-transformation where applicable.",
  align   = c("l","l","r","r","r","r","r","r")
)
```

```{r log-comparison-bar, fig.height=4.0, fig.cap="**Figure 7.** Side-by-side comparison of R², RMSE, RPD, and CCC for the best model under each log-mode strategy (both: SG 1st derivative). Differences are small, with the no-log mode showing a marginal advantage in RMSE and RPD."}
modes_df <- data.frame(
  mode  = c("Log₁p(SOC)", "Original SOC"),
  r2    = c(m_comp$r2[m_comp$log_mode == "log"],    m_comp$r2[m_comp$log_mode == "no_log"]),
  rmse  = c(m_comp$rmse[m_comp$log_mode == "log"],  m_comp$rmse[m_comp$log_mode == "no_log"]),
  rpd   = c(m_comp$rpd[m_comp$log_mode == "log"],   m_comp$rpd[m_comp$log_mode == "no_log"]),
  ccc   = c(m_comp$ccc[m_comp$log_mode == "log"],   m_comp$ccc[m_comp$log_mode == "no_log"])
)

modes_long <- tidyr::pivot_longer(modes_df, -mode, names_to = "metric", values_to = "value")
modes_long$metric <- factor(modes_long$metric,
  levels = c("r2","rmse","rpd","ccc"),
  labels = c("R²","RMSE (%)","RPD","CCC"))
modes_long$mode <- factor(modes_long$mode, levels = c("Log₁p(SOC)", "Original SOC"))

ggplot(modes_long, aes(x = mode, y = value, fill = mode)) +
  geom_col(width = 0.55, show.legend = FALSE) +
  geom_text(aes(label = round(value, 2)), vjust = -0.4, size = 3.8, fontface = "bold") +
  scale_fill_manual(values = c("Log₁p(SOC)" = "#2980b9", "Original SOC" = "#27ae60")) +
  facet_wrap(~metric, scales = "free_y", nrow = 1) +
  labs(x = NULL, y = NULL,
       title = "Log mode comparison — best model (SG 1st Derivative)") +
  theme_bw(base_size = 11) +
  theme(
    strip.background  = element_rect(fill = "#2c3e50"),
    strip.text        = element_text(colour = "white", face = "bold"),
    panel.grid.major.x = element_blank(),
    plot.title        = element_text(face = "bold", size = 11)
  )
```

<div class="callout-green">
**Finding.** Both log-mode strategies select **SG 1st derivative** as the best preprocessing treatment. The difference in predictive performance is **small and practically negligible**: R² differs by `r abs(round(m_comp$r2[1] - m_comp$r2[2], 2))` and RMSE differs by `r abs(round(m_comp$rmse[1] - m_comp$rmse[2], 2))` percentage points. The no-log model shows a slight advantage in RMSE and RPD, likely because SOC in the OSSL sample covers a wide range where the log transform compresses extreme values and the RF can handle this directly. For highly skewed BMP project targets (TP, WEP), the log transformation is likely to provide greater benefit.
</div>

---

# Variable Importance

Variable importance (mean impurity decrease, averaged across 30 CV resamples) identifies the spectral wavelengths most informative for SOC prediction under each preprocessing strategy.

```{r importance-plot, fig.height=4.5, fig.cap="**Figure 8.** Spectral variable importance profiles for SG 1st derivative (no-log mode, top) and SNV (no-log mode, bottom). Importance is averaged over 30 CV resamples and normalised to [0, 1]. Key absorption features associated with SOC in the 1400–1800 nm (C-H, O-H overtones) and 2000–2300 nm (C=O, CH₂ combination bands) regions are commonly highlighted."}
knitr::include_graphics(c(
  "output/figures/importance_oc_usda.c729_w.pct_raw_sg_deriv1.png",
  "output/figures/importance_oc_usda.c729_w.pct_raw_snv.png"
))
```

---

# MSD Error Decomposition

The Mean Squared Deviation (MSD) decomposes model error into three additive components (Gauch et al. 2003):

$$\text{MSE} = \underbrace{(\bar{\hat{y}} - \bar{y})^2}_{\text{SB (Squared Bias)}} + \underbrace{(1-b)^2 \sigma_{\hat{y}}^2}_{\text{NU (Non-Unity slope)}} + \underbrace{(1-r^2)\sigma_y^2}_{\text{LC (Lack of Correlation)}}$$

A well-calibrated model has SB ≈ 0 (no systematic offset) and NU ≈ 0 (predictions correctly scaled), with most error in LC (irreducible noise).

```{r msd-table}
msd_df <- dplyr::bind_rows(
  m_log %>% dplyr::mutate(Mode = "Log₁p"),
  m_raw %>% dplyr::mutate(Mode = "Original")
) %>%
  dplyr::mutate(
    Preprocessing = prep_labels[preprocessing],
    MSE           = round(rmse^2, 2),
    `SB (%)`      = round(sb / rmse^2 * 100, 1),
    `NU (%)`      = round(nu / rmse^2 * 100, 1),
    `LC (%)`      = round(lc / rmse^2 * 100, 1)
  ) %>%
  dplyr::select(Mode, Preprocessing, MSE, `SB (%)`, `NU (%)`, `LC (%)`)

knitr::kable(
  msd_df,
  caption = "**Table 4.** MSD decomposition for all treatment × log-mode combinations. SB = squared bias; NU = non-unity slope (scale error); LC = lack of correlation (random error). Values in % of total MSE.",
  align   = c("l","l","r","r","r","r")
)
```

<div class="callout-orange">
**Interpretation.** Across all treatments and log modes, the largest error component is **LC** (lack of correlation, irreducible noise), which is expected and desirable — it means the RF is not introducing systematic biases. The **SB** component is consistently small for the no-log mode (near-zero mean error), while the log mode shows slightly higher SB due to back-transformation of Jensen's inequality effects. **NU** is small across all cases, indicating predictions are correctly scaled (slope ≈ 1).
</div>

---

# Best Model Summary

```{r best-summary-table}
data.frame(
  Setting = c(
    "Sensor", "Wavelength range", "N wavelengths (predictors)",
    "Sample size (n)", "Unique soil layers",
    "Preprocessing", "Log transformation",
    "num.trees", "mtry", "min.node.size",
    "CV strategy", "Total resamples",
    "R² (mean ± SD)", "RMSE (mean ± SD)", "MAE (mean ± SD)",
    "RPD (mean ± SD)", "RPIQ (mean ± SD)", "CCC (mean ± SD)",
    "ME (mean ± SD) — bias"
  ),
  Value = c(
    "ASD FieldSpec 4",
    sprintf("%d–%d nm", WAVE_MIN, WAVE_MAX),
    sprintf("%d", n_wave),
    sprintf("%d", n_final),
    sprintf("%d", n_uuid_final),
    "SG 1st Derivative (m=1, w=11, p=3) applied to SG-smoothed base",
    "None (original SOC % scale)",
    "Selected by OOB RMSE (192-combination 3D grid search)",
    "Selected by OOB RMSE",
    "Selected by OOB RMSE",
    "10-fold × 3-repeat stratified grouped CV",
    "30",
    fmt_ms(best_row$r2,   best_row$r2_sd),
    sprintf("%s %%", fmt_ms(best_row$rmse, best_row$rmse_sd)),
    sprintf("%s %%", fmt_ms(best_row$mae,  best_row$mae_sd)),
    fmt_ms(best_row$rpd,  best_row$rpd_sd),
    fmt_ms(best_row$rpiq, best_row$rpiq_sd),
    fmt_ms(best_row$ccc,  best_row$ccc_sd),
    sprintf("%s %%", fmt_ms(best_row$me,   best_row$me_sd))
  )
) |>
  knitr::kable(
    caption = "**Table 5.** Summary of the best-performing model configuration (SG 1st Derivative, no-log scale).",
    col.names = c("Setting", "Value"),
    align = c("l", "l")
  )
```

---

# Conclusions

1. **SG 1st derivative is consistently the best preprocessing treatment** across all metric dimensions and both log modes. It removes additive baseline offsets and multiplicative scatter simultaneously, isolating spectral slope information that is strongly coupled to SOC absorption features.

2. **Log transformation provides no meaningful benefit** for this OSSL prototype dataset (ΔRMSE < 0.3 %). For the BMP project dataset, this should be re-evaluated property by property — highly right-skewed targets such as TP, WEP, and Fe are more likely to benefit from the log₁p transformation.

3. **Model quality is good to excellent** under the RPD classification of Chang et al. (2001): RPD = `r round(best_row$rpd, 2)` (> 3.0) places the best model in the "excellent" category. The CCC of `r round(best_row$ccc, 2)` confirms both precision and accuracy (minimal systematic bias).

4. **Spectral replicates were preserved without averaging**, with leakage prevented via grouped fold assignment by layer UUID (Kaufman & Rosset 2012). This approach avoids the bias introduced by retaining replicates across fold boundaries while preserving the full spectral information available.

5. **Next steps:** Replace OSSL prototype data with the BMP Years 1–5 spectral-lab dataset. Extend the target list to include pH, TP, M3-P, WEP, TC, TN, Fe, Al, Ca, and Mg. Validate whether the OSSL-calibrated model transfers to the BMP site-specific samples.

---

# References

- Barnes, R.J., Dhanoa, M.S., & Lister, S.J. (1989). Standard normal variate transformation and de-trending of near-infrared diffuse reflectance spectra. *Applied Spectroscopy*, 43(5), 772–777. <https://doi.org/10.1366/0003702894202201>

- Bellon-Maurel, V., Fernandez-Ahumada, E., Palagos, B., Roger, J.-M., & McBratney, A. (2010). Critical review of chemometric indicators commonly used for assessing the quality of the prediction of soil attributes by NIR spectroscopy. *TrAC Trends in Analytical Chemistry*, 29(9), 1073–1081. <https://doi.org/10.1016/j.trac.2010.05.006>

- Breiman, L. (2001). Random forests. *Machine Learning*, 45(1), 5–32. <https://doi.org/10.1023/A:1010933404324>

- Chang, C.-W., Laird, D.A., Mausbach, M.J., & Hurburgh, C.R. (2001). Near-infrared reflectance spectroscopy–principal components regression analyses of soil properties. *Soil Science Society of America Journal*, 65(2), 480–490. <https://doi.org/10.2136/sssaj2001.652480x>

- Clingensmith, C.M. & Grunwald, S. (2022). Predicting soil properties and interpreting Vis-NIR models from across continental United States. *Sensors*, 22(9), 3187. <https://doi.org/10.3390/s22093187>

- Gauch, H.G., Hwang, J.T.G., & Fick, G.W. (2003). Model evaluation by comparison of model-based predictions and measured values. *Agronomy Journal*, 95(6), 1442–1446. <https://doi.org/10.2134/agronj2003.1442>

- Kaufman, S. & Rosset, S. (2012). Leakage in data mining: Formulation, detection, and avoidance. *ACM Transactions on Knowledge Discovery from Data*, 6(4), 15. <https://doi.org/10.1145/2382577.2382579>

- Liaw, A. & Wiener, M. (2002). Classification and regression by randomForest. *R News*, 2(3), 18–22. <https://cran.r-project.org/doc/Rnews/>

- Lin, L.I.-K. (1989). A concordance correlation coefficient to evaluate reproducibility. *Biometrics*, 45(1), 255–268. <https://doi.org/10.2307/2532051>

- Probst, P., Wright, M.N., & Boulesteix, A.-L. (2019). Hyperparameters and tuning strategies for random forest. *WIREs Data Mining and Knowledge Discovery*, 9(3), e1301. <https://doi.org/10.1002/widm.1301>

- Rinnan, Å., van den Berg, F., & Engelsen, S.B. (2009). Review of the most common pre-processing techniques for near-infrared spectra. *TrAC Trends in Analytical Chemistry*, 28(10), 1201–1222. <https://doi.org/10.1016/j.trac.2009.07.007>

- Safanelli, J.L., Hengl, T., Parente, L., et al. (2023). Open Soil Spectral Library (OSSL): Building reproducible soil calibration models through open development and community engagement. *PLOS ONE*. <https://doi.org/10.1371/journal.pone.0296545>

- Savitzky, A. & Golay, M.J.E. (1964). Smoothing and differentiation of data by simplified least squares procedures. *Analytical Chemistry*, 36(8), 1627–1639. <https://doi.org/10.1021/ac60214a047>

- Viscarra Rossel, R.A. & Behrens, T. (2010). Using data mining to model and interpret soil diffuse reflectance spectra. *Geoderma*, 158(1–2), 46–54. <https://doi.org/10.1016/j.geoderma.2009.12.025>

---

<div style="margin-top:40px; padding:16px; background:#f8f9fa; border-radius:6px; font-size:0.85em; color:#7f8c8d; text-align:center;">
Report generated with R · ranger · OSSL — BMP Project, University of Florida<br>
© 2026 Hugo Machado Rodrigues ·
<a href="https://orcid.org/0000-0002-8070-8126">ORCID</a> ·
<a href="https://github.com/HugoMachadoRodrigues">GitHub</a> ·
<a href="https://www.researchgate.net/profile/Hugo-Rodrigues-12">ResearchGate</a><br>
<em>Pipeline code: <a href="https://github.com/HugoMachadoRodrigues/soil-rf-spectral">github.com/HugoMachadoRodrigues/soil-rf-spectral</a></em>
</div>
